---
title: "*t*-Test in R"
author: "Md. Ahsanul Islam"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


# Required Packages

```{r}
library(ggplot2)
library(reshape2)
library(car)
library(ggpubr)
```

# One sample t-test

## Manual Calculation

```{r}
set.seed(123)
sample_data <- rnorm(30, mean = 50, sd = 10)
mean(sample_data)
sd(sample_data)
```

Hypothesis:\
$H_0: \mu = \mu_0$\
$H_1: \mu \neq \mu_0$\

```{r}
# mu_0 = 55
SE <- sd(sample_data)/sqrt(length(sample_data))
t_cal <- (mean(sample_data) - 55)/SE
print(t_cal)
t_crit <- qt(0.05/2, df = length(sample_data)-1, lower.tail = TRUE)
t_crit

t_cal <= t_crit  # decision: reject null
pt(t_cal, df = 29, lower.tail = TRUE) + pt(-t_cal, df = 29, lower.tail = FALSE)
```

Calculating confidence interval:
```{r}
mean(sample_data) + c(-1, 1) * abs(t_crit) * SE
```

## Using Function

Generate sample data:
```{r}
set.seed(123)
sample_data <- rnorm(30, mean = 50, sd = 10)
```

Testing normality using Shapiro-Wilk test:
```{r}
shapiro.test(sample_data)
```

$H_0:$ Data follows normal distribution.   
$H_1:$ Data does not follow normal distribution.   

Since the p-value is greater than the level of significance ($\alpha$ = 0.05), we do not have enough statistical evidence to reject the null hypothesis.

One-sample t-test:

Hypothesis:\
$H_0: \mu = 50$\
$H_1: \mu \neq 50$\

Using function, perform the two tailed t-test:
```{r}
t.test(sample_data, mu = 53, conf.level = 0.95)
```

If p-value < 0.05, then reject null.
Decision is not rejected.

```{r}
t.test(sample_data, mu = 53, conf.level = 0.99)
```

If p-value < 0.01, then reject null. 
Decision is not rejected.

```{r}
t.test(sample_data, mu = 53, conf.level = 0.90)
```

If p-value < 0.10, then reject null. 
Decision is rejected.


Hypothesis:\
$H_0: \mu <= 40$\
$H_1: \mu > 40$\
```{r}
t.test(sample_data, mu = 40, alternative = "greater")
```

Hypothesis:\
$H_0: \mu >= 58$\
$H_1: \mu < 58$\
```{r}
t.test(sample_data, mu = 58, alternative = "less")
```


```{r}
ggplot(data.frame(Value = sample_data), aes(x = Value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, fill = "blue", alpha = 0.5) +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Sample Data Distribution", x = "Value", y = "Density") +
  theme_minimal()
```

# Two-samples t-test

Use built-in dataset: mtcars (comparing mpg for automatic vs manual cars):
```{r}
data(mtcars)
```

Split into two groups based on transmission type":
```{r}
auto_mpg <- mtcars$mpg[mtcars$am == 0]    # Automatic
manual_mpg <- mtcars$mpg[mtcars$am == 1]  # Manual
```

$H_0$: Automatic cars and manual cars have equal average mpg.   
$H_1$: Automatic cars and manual cars have unequal average mpg.   

```{r}
mean(auto_mpg)
mean(manual_mpg)
```

```{r}
var(auto_mpg)
var(manual_mpg)
```



Normality test for both groups:
```{r}
shapiro.test(auto_mpg)
shapiro.test(manual_mpg)
```

Check variance homogeneity (Levene's test):
```{r}
leveneTest(mpg ~ factor(am), data = mtcars, center = "mean")
```

Perform two-sample t-test:
```{r}
t.test(auto_mpg, manual_mpg, var.equal = FALSE) 
```


```{r}
# Visualize the data
ggplot(mtcars, aes(x = factor(am), y = mpg, fill = factor(am))) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  labs(title = "MPG Comparison: Automatic vs Manual", 
       x = "Transmission (0 = Auto, 1 = Manual)", 
       y = "Miles Per Gallon") +
  scale_fill_manual(values = c("blue", "red"), 
                    labels = c("Automatic", "Manual")) +
  theme_minimal()
```


# Paired samples t-test

Generate 30 observations:
```{r}
set.seed(123)
before <- round(rnorm(30, mean = 300, sd = 10), 0)
summary(before)
# after a course
after <- before + round(rnorm(30, mean = 5, sd = 5), 0) # Simulating a increase
summary(after)
```

```{r}
df <- data.frame(
  ID = 1:30,
  Before = before,
  After = after
)
```

```{r}
mean(after) - mean(before)
```


$H_0:$ Before and after the course the true GRE average score of the students stays the same.   
$H_1:$ Before and after the course the true GRE average score of the students does not remain the same.   

Perform Paired t-test
```{r}
t.test(x = after, y = before, paired = TRUE, alternative = "two.sided")
```

```{r}
# Visualize the differences
df <- data.frame(
  ID = 1:30,
  Before = before, 
  After = after
  )
df_long <- melt(df, id.vars = "ID")
ggplot(df_long, aes(x = variable, y = value, group = ID)) +
  geom_point(aes(color = variable), size = 3) +
  geom_line() +
  labs(title = "Paired Samples (Before vs. After)", 
       y = "Values", x = "Condition") +
  theme_minimal()

ggpaired(df_long, 
         x = "variable", 
         y = "value",
         color = "variable", 
         line.color = "gray", 
         line.size = 0.4,
         palette = "jco") +
  stat_compare_means(paired = TRUE, method = "t.test")
```

