---
title: "GLM and Logistic Regression"
author: "Md. Ahsanul Islam"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = NA)
```

# Packages

```{r}
library(tidyverse)
library(titanic)
```


# Data

```{r}
data <- titanic_train

data <- data %>%
  select(Survived, Pclass, Sex, Age, Fare) %>%  
  filter(!is.na(Age)) %>%  
  mutate(Survived = factor(Survived, levels = c(0, 1)),
         Sex = factor(Sex, levels = c("male", "female")))
```

```{r}
data %>% 
  group_by(Sex, Pclass) %>% 
  summarise(mean(Age), n())

mean(data$Age)
```


```{r}
model <- glm(Survived ~ Pclass + Sex + Age, data = data, family = binomial)
summary(model)
coef(model)
exp(coef(model))
exp(coef(model)[-1])  # odds ratio
```

Each increase in class decreases survival odds by 72% (since odds = 0.28).    
Females were 92% more likely to survive than males.     
Males are 92% less likely to survive than females.    
Each year increase in age decreases odds of survival by 4%.   
Fare has no meaningful effect.     

# Prediction

Predict survival probabilities for new data:

```{r}
new_data <- data.frame(
  Pclass = c(1, 2, 3),  
  Sex = factor(c("male", "female", "female"), levels = c("male", "female")),
  Age = c(30, 25, 40)
)

new_data$link <- predict(model, newdata = new_data, type = "link") # log(odds)
new_data

new_data$Predicted_Prob <- predict(model, newdata = new_data, type = "response")
new_data

new_data$terms <- predict(model, newdata = new_data, type = "terms")
new_data
```


```{r}
test_dat <- titanic_test

test_dat$Predicted <- predict(model, test_dat, type = "response")
test_dat$Survived <- ifelse(test_dat$Predicted >= 0.5, "Yes", "No")
```


# Evaluation Metrices

Predict survival using the model (threshold = 0.5):
```{r}
predicted_class <- ifelse(predict(model, type = "response") >= 0.5, 1, 0)  # Convert probability to class
```

## Confusion metrix:

```{r}
caret::confusionMatrix(factor(predicted_class), data$Survived)
caret::confusionMatrix(factor(predicted_class), data$Survived, mode = "prec_recall")
```

## ROC Curve

```{r}
library(pROC)

predicted_prob <- predict(model, type = "response")
roc_curve <- roc(data$Survived, predicted_prob)

plot(roc_curve, col = "blue", main = "ROC Curve for Titanic Survival Model")

auc_value <- auc(roc_curve)
print(paste("AUC:", auc_value))
```

## McFadden's R-squared

```{r}
null_model <- glm(Survived ~ 1, data = data, family = binomial)  # Null model (only intercept, no predictors)
R2 <- 1 - (logLik(model) / logLik(null_model))
print(paste("McFadden's R-squared:", R2 |> round(3)))
```

R squared > 0.2: Good model.
R squared > 0.4: Strong model.


