---
title: "ANOVA and Post-hoc analysis in R"
author: "Md. Ahsanul Islam"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = NA)
```

# Packages

```{r}
library(dplyr)
library(ggplot2)
```

# Data

```{r}
set.seed(42)  # for reproducibility

groupA <- rnorm(30, mean = 50, sd = 5)
groupB <- rnorm(30, mean = 55, sd = 5) 
groupC <- rnorm(30, mean = 60, sd = 5) 
groupD <- rnorm(30, mean = 47, sd = 5)

my_data <- data.frame(
  value = c(groupA, groupB, groupC, groupD),
  group = factor(rep(c("A", "B", "C", "D"), each = 30))
)
```



```{r}
my_data %>% 
  group_by(group) %>% 
  summarize(Avg = mean(value))

my_data %>% 
  ggplot(aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.6) +
  facet_wrap(vars(group), ncol = 1)
```


# Assumptions Test

Normality test:   

Null hypo: Data does not deviate from normal distribution.   
Alt hypo: Data deviates from normal distribution.    


```{r}
shapiro.test(groupA) # null not rejected
shapiro.test(groupB) # null not rejected
shapiro.test(groupC) # null not rejected
```

Homogeneity of variance test:

Null: Group variances are equal.
Alt: At least one group variance differ than others.

```{r}
car::leveneTest(value ~ group, data = my_data)
```


# ANOVA

$$
H_0: \mu_1 = \mu_2 =\mu_3 \newline
$$

$H_1:$ At least one of the group mean is not equal to others.


```{r}
anova_model <- aov(value ~ group, data = my_data)
summary(anova_model)
```

# Post-Hoc


The following procedure is wrong:
```{r}
my_data_sub <- my_data %>% 
  filter(group %in% c("A","B"))
t.test(value ~ group, my_data_sub, var.equal = TRUE)

my_data_sub <- my_data %>% 
  filter(group %in% c("A","C"))
t.test(value ~ group, my_data_sub, var.equal = TRUE)

my_data_sub <- my_data %>% 
  filter(group %in% c("A","D"))
t.test(value ~ group, my_data_sub, var.equal = TRUE)
```

We need to use post hoc tests: 

```{r}
TukeyHSD(anova_model, "group", conf.level = 1-0.05)
```

```{r}
pairwise.t.test(my_data$value, my_data$group, p.adjust.method = "bonferroni")
```

```{r}
DescTools::ScheffeTest(anova_model, "group", conf.level = 0.95)
```



# Example


```{r}
data <- readxl::read_excel("Data/StudentSurveyData.xlsx")
str(data)
data <- data %>% 
  mutate(Computer = factor(Computer))
```


```{r}
summary(data %>% select(Computer, GPA))
```

```{r}
data %>% 
  group_by(Computer) %>% 
  summarize(mean(GPA))

data %>% 
  ggplot(aes(x = GPA)) + 
  geom_histogram() +
  facet_wrap(~Computer, ncol = 1)
```


```{r}
anova_model_stu <- aov(GPA ~ Computer, data)
summary(anova_model_stu)
```

```{r}
TukeyHSD(anova_model_stu)
```



```{r}
data %>% 
  group_by(Employment) %>% 
  summarize(mean(Spending))

data %>% 
  ggplot(aes(x = Spending)) + 
  geom_boxplot() +
  facet_wrap(~Employment, ncol = 1)
```


```{r}
anova_model_stu2 <- aov(Spending ~ Employment, data)
summary(anova_model_stu2)

aov(lm(Spending ~ Employment, data))
```

```{r}
TukeyHSD(anova_model_stu2)
```
